Experimental and numerical studies of the effects of resonant and random excitations of the proton beam in the LHC, with applications to the design of a pulsed hollow electron lens for active halo control

### 2018-03-22 resex.tex
Minor changes in layout and fonts. Fixed the rest of the ordinal numbers.

\documentclass[
prstab,reprint
%,draft
%,preprint,tightenlines
,linenumbers
,longbibliography
,preprintnumbers
,showpacs
,showkeys
,amsfonts,amssymb,amsmath
,floatfix
]{revtex4-1}

\usepackage{bm}
\usepackage{dcolumn}
\usepackage{graphicx}
\usepackage{hyperref}

@@

\begin{table*}
  \caption{Beam parameters and machine configuration for the two
    resonant excitation experiments of~2016
    and~2017~\cite{resexmd2016, resexmd2017}. The plane of the
    excitation is abbreviated as H for horizontal, V for vertical and
    H+V for horizontal and vertical at the same time.}
  \label{tab:md_param}
  \begin{ruledtabular}
    \begin{tabular}{lcc}
      Parameter & Experiment 2016 & Experiment 2017  \\
      \colrule
      beam &\multicolumn{2}{c}{Beam~1} \\
      beam energy &\multicolumn{2}{c}{injection energy, 450 GeV} \\\hline
      single bunch intensity &\multicolumn{2}{c}{$0.7\times10^{11}$} \\
      normalized emittance &\multicolumn{2}{c}{$2.5$--\q{3.5}{\mu m}} \\
      $4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
      $1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
      number of bunches & $12 \times 4 = 48$ & $3 \times 72 = 216$ \\
                &  & (+ 1 pilot + 12 nominal) \\\hline
      injection optics, $\beta^* = \q{11}{m}$ & standard optics 2016 & standard optics 2017\\
      Landau-damping octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MO}}
                                  = \q{+19.6}{A}$ for MOF circuit and}\\
                & \multicolumn{2}{c}{$I_{\mathrm{MO}} = \q{-19.6}{A}$
                  for MOD circuit (standard 2016 settings)} \\ \hline
      working point $(Q_x, Q_y)$ & (64.28, 59.31) & (62.27, 60.295) \\
      chromaticity $(Q'_x, Q'_y)$ & \multicolumn{2}{c}{(+15, +15)}\\ \hline
      pulsing patterns  & 7th~turn H
                                  &7th~turn H, V, H+V \\
                & $10^{\mathrm{th}}$~turn V & $8^{\mathrm{th}}$~turn
                                              H, V, H+V \\
                & &  random  H, V, H+V\\
    \end{tabular}
  \end{ruledtabular}
\end{table*}

@@

\begin{figure*}
  \includegraphics[width=0.9\textwidth]{bunchfilling_measured.png}	
  \caption{Example of bunch-by-bunch beam centroid motion for Beam~1
    in the horizontal (B1H, left) and vertical (B1V, right) planes,
    detected by the real-time transverse activity monitor during the
    2017 experiment, when the beam was excited every
    $8^\mathrm{th}$~turn.}
  \label{fig:fill_meas} 
\end{figure*}

@@

The choice of excitation patterns for the experiments was guided by
losses and emittance growths calculated in numerical tracking
simulations, described below. It was chosen to study experimentally
the two pulsing patterns with the largest calculated effects on the
beam (7th- and $10^{\mathrm{th}}$-turn pulsing), one pattern with no
effect ($8^{\mathrm{th}}$-turn pulsing), and the random excitation. In
order to quantify the reproducibility of the results under different
machine configurations, one pulsing pattern (\seventhtp) was tried
first in~2016 and then repeated in~2017.

@@

Figure~\ref{fig:fill_meas} shows an example of the oscillation
amplitude of each bunch centroid in both horizontal and vertical
planes, captured by the LHC real-time transverse activity monitor
during the 2017 experiments, when the beam was excited every
$8^{\mathrm{th}}$~turn. The measured activity shows the typical
amplitude of bunch centroid excursions and reflects the expected
excitation pattern.

@@

In this Section, we present the experimental observations and compare
them with the results of numerical simulations.  In
Section~\ref{sec:pattern}, we show which pulsing patterns are
predicted to be the most efficient in the LHC.  Next, we present
results for each of the specific excitation patterns that could be
tested experimentally, namely pulsing every $10^{\mathrm{th}}$~turn
(Section~\ref{sec:simex10}), 7th~turn
(Section~\ref{sec:simex7}), $8^{\mathrm{th}}$~turn
(Section~\ref{sec:simex8}), and the random excitation
(Section~\ref{sec:simexran}).  In Section~\ref{sec:damp}, we describe
how the transverse damping system influenced the effects of the
external excitation sources on losses, beam distributions, and
emittances.

@@

\begin{figure*}
  \begin{tabular}{cc}
    no excitation & $10^{\mathrm{th}}$, H+V \\
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png} &
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
    \\
    7th, H & 7th, V \\
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png} &
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png} \\
  \end{tabular}
  \caption{Frequency-map analysis in betatron tune space based on the
    2016 injection optics with no machine errors and tunes (64.28,
    59.31): without excitation (top left); $10^{\mathrm{th}}$~turn
    pulsing in both horizontal and vertical planes (top left);
    \seventhtp\ in the horizontal (bottom left) and vertical (bottom
    right) planes. The excitation amplitude is 120~nrad in the
    corresponding plane. The colors (blue to red) represent the tune
    jitter of tracked particles starting at each given location in
    tune space~\cite{fmalaskar}. The absence of a strong excitation of
    any resonance in case of vertical \seventhtp\ and the strong
    excitation in case of horizontal pulsing confirms the excitation
    of the $7 Q_x$ resonance. For $10^{\mathrm{th}}$~turn pulsing,
    there is in contrast no significant difference between H, V, or
    H+V (Fig.~\ref{fig:fma:10}).}
  \label{fig:patternfma}
\end{figure*}

@@

As an example, the simulation results for the 2016 experiment are
shown in Fig.~\ref{fig:patternsim}.  A clear dependence of both losses
and emittances on the pattern can be seen.  The largest losses are
predicted for $3^{\mathrm{rd}}$, 7th and
$10^{\mathrm{th}}$~turn pulsing, and for a uniform random
excitation. Because the bunch length decreases with the number of lost
particles, losses are attributed to off-momentum particles hitting the
transverse aperture. Significant emittance growth is visible only for
7th- and \tenthtp\ and for the
random excitation. Compared to any resonant excitation, the random
excitation shows by far the strongest effect.

The effects of 7th- and \tenthtp\ are also observed in the absence of
machine lattice errors. In this case, the only sources of nonlinearity
are sextupoles and octupoles~\cite{md_sim_hel_res_ex_fitterer},
suggesting that these nonlinearities are responsible for the
pronounced beam sensitivity. The driven resonances are revealed by the
frequency-map analysis, shown in Fig.~\ref{fig:patternfma}. The
$7 Q_x$ resonance is excited by the \seventhtp, and the $10 Q_x$ and
$10 Q_y$ resonances by the $10^{\mathrm{th}}$~turn pulsing. As
octupoles can only drive even resonances, the sources of the $7 Q_x$
resonances are the sextupoles, while the octupoles generate the tune
footprint. The other pulsing patterns do not exhibit any increase in
losses or emittance growth without magnetic errors. Their effect can
be thus attributed to an interaction of the excitation with the
lattice errors, implying also a sensitivity to the chosen random seed
in simulations and to the specific lattice in experiments.

@@

\subsection{Pulsing every $10^{\mathrm{th}}$ turn}
\label{sec:simex10}

\begin{figure*}
  \begin{tabular}{ccc}
    \includegraphics[width=\thirdwidth]{2016_bunch_intensity_v10th_no_damper_avg.png}&
    \includegraphics[width=\thirdwidth]{2016_emith_avg_rel_v10th_no_damper.png}&
    \includegraphics[width=\thirdwidth]{2016_emitv_avg_rel_v10th_no_damper.png}\\
  \end{tabular}
  \caption{Summary of the 2016 experiments on $10^{\mathrm{th}}$~turn
    pulsing in the vertical plane: losses (left), horizontal
    emittances (middle) and vertical emittances (right), relative to
    their initial values. The transverse damping system was not active
    in this case. The 3 excitation periods are labeled in black
    according to the value of the maximum excitation amplitude
    $A_{\mathrm{max}} = 5 \, \Delta A$: no excitation, 48~nrad or
    96~nrad. The 4~bunches experiencing the same excitation amplitude
    $n \, \Delta A$ ($n = 0, \ldots, 5$) are grouped by color. The
    data are averaged over the 4~bunches, with the envelope
    representing the standard deviation. The area with a blue
    background highlights qualitatively the fast adjustment period of
    the beam distribution, transitioning into a new equilibrium state
    (indicated by the gray background).}
  \label{fig:10thexp}
\end{figure*}

@@

An insight on how the change in tune entails a change in driving
resonances is provided by frequency-map analysis.
Figure~\ref{fig:patternfma} (bottom left and bottom right plots),
which was described previously, shows a strong increase in diffusion
for \seventhtp\ in the horizontal plane around the $7Q_x$ resonances,
and only small changes for pulsing in the vertical plane. In the case
of the 2017 optics and tune, both the $7Q_x$ and the $7Q_y$ resonances
cross the tune footprint (Fig.~\ref{fig:7th2017fma}, top left). Both
of these resonances are excited by \seventhtp, as one can see from the
increase in tune jitter around the corresponding resonance lines for
each excitation plane separately, and thus also for their combined
action (Fig.~\ref{fig:7th2017fma}, bottom left, bottom right, and top
right plots). Although much weaker, an increase in tune diffusion is
also observed around the $7Q_x + 7Q_y$ line when pulsing in the
vertical plane or in both planes (diagonal yellow-green line in
Fig.~\ref{fig:7th2017fma}, top right and bottom right plots). This
suggests an excitation of the $14^{\mathrm{th}}$-order resonance. The
reason why this phenomenon is much less pronounced for horizontal
excitations is not understood.

>>>>

\documentclass[
prstab
,reprint
%,draft
%,preprint,tightenlines
,linenumbers
,longbibliography
,preprintnumbers
,showpacs
,showkeys
,amsfonts,amssymb,amsmath
,floatfix
]{revtex4-1}

\usepackage{bm}
\usepackage{dcolumn}
\usepackage{graphicx}
\usepackage[colorlinks=true, linkcolor=blue, citecolor=blue,
            filecolor=blue, urlcolor=blue]{hyperref}

\usepackage{mathptmx}
\linespread{1.1}            
\usepackage[T1]{fontenc}

@@

\begin{table*}
  \caption{Beam parameters and machine configuration for the two
    resonant excitation experiments of~2016
    and~2017~\cite{resexmd2016, resexmd2017}. The plane of the
    excitation is abbreviated as H for horizontal, V for vertical and
    H+V for horizontal and vertical at the same time.}
  \label{tab:md_param}
  \begin{ruledtabular}
    \begin{tabular}{lcc}
      Parameter & Experiment 2016 & Experiment 2017  \\
      \colrule
      beam &\multicolumn{2}{c}{Beam~1} \\
      beam energy &\multicolumn{2}{c}{injection energy, 450 GeV} \\\hline
      single bunch intensity &\multicolumn{2}{c}{$0.7\times10^{11}$} \\
      normalized emittance &\multicolumn{2}{c}{$2.5$--\q{3.5}{\mu m}} \\
      $4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
      $1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
      number of bunches & $12 \times 4 = 48$ & $3 \times 72 = 216$ \\
                &  & (+ 1 pilot + 12 nominal) \\\hline
      injection optics, $\beta^* = \q{11}{m}$ & standard optics 2016 & standard optics 2017\\
      Landau-damping octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MO}}
                                  = \q{+19.6}{A}$ for MOF circuit and}\\
                & \multicolumn{2}{c}{$I_{\mathrm{MO}} = \q{-19.6}{A}$
                  for MOD circuit (standard 2016 settings)} \\ \hline
      working point $(Q_x, Q_y)$ & (64.28, 59.31) & (62.27, 60.295) \\
      chromaticity $(Q'_x, Q'_y)$ & \multicolumn{2}{c}{(+15, +15)}\\ \hline
      pulsing patterns  & 7th~turn H & 7th~turn H, V, H+V \\
                & 10th~turn V & 8th~turn H, V, H+V \\
                & &  random  H, V, H+V\\
    \end{tabular}
  \end{ruledtabular}
\end{table*}

@@

\begin{figure*}
  \includegraphics[width=0.9\textwidth]{bunchfilling_measured.png}	
  \caption{Example of bunch-by-bunch beam centroid motion for Beam~1
    in the horizontal (B1H, left) and vertical (B1V, right) planes,
    detected by the real-time transverse activity monitor during the
    2017 experiment, when the beam was excited every 8th~turn.}
  \label{fig:fill_meas} 
\end{figure*}

@@

The choice of excitation patterns for the experiments was guided by
losses and emittance growths calculated in numerical tracking
simulations, described below. It was chosen to study experimentally
the two pulsing patterns with the largest calculated effects on the
beam (7th- and \tenthtp), one pattern with no effect (\eighthtp), and
the random excitation. In order to quantify the reproducibility of the
results under different machine configurations, one pulsing pattern
(\seventhtp) was tried first in~2016 and then repeated in~2017.

@@

Figure~\ref{fig:fill_meas} shows an example of the oscillation
amplitude of each bunch centroid in both horizontal and vertical
planes, captured by the LHC real-time transverse activity monitor
during the 2017 experiments, when the beam was excited every
8th~turn. The measured activity shows the typical amplitude of bunch
centroid excursions and reflects the expected excitation pattern.

@@

In this Section, we present the experimental observations and compare
them with the results of numerical simulations.  In
Section~\ref{sec:pattern}, we show which pulsing patterns are
predicted to be the most efficient in the LHC.  Next, we present
results for each of the specific excitation patterns that could be
tested experimentally, namely pulsing every 10th~turn
(Section~\ref{sec:simex10}), 7th~turn (Section~\ref{sec:simex7}),
8th~turn (Section~\ref{sec:simex8}), and the random excitation
(Section~\ref{sec:simexran}).  In Section~\ref{sec:damp}, we describe
how the transverse damping system influenced the effects of the
external excitation sources on losses, beam distributions, and
emittances.

@@

\begin{figure*}
  \begin{tabular}{cc}
    No excitation & 10th, H+V \\
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png} &
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
    \\
    7th, H & 7th, V \\
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png} &
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png} \\
  \end{tabular}
  \caption{Frequency-map analysis in betatron tune space based on the
    2016 injection optics with no machine errors and tunes (64.28,
    59.31): without excitation (top left); \tenthtp\ in both
    horizontal and vertical planes (top left); \seventhtp\ in the
    horizontal (bottom left) and vertical (bottom right) planes. The
    excitation amplitude is 120~nrad in the corresponding plane. The
    colors (blue to red) represent the tune jitter of tracked
    particles starting at each given location in tune
    space~\cite{fmalaskar}. The absence of a strong excitation of any
    resonance in case of vertical \seventhtp\ and the strong
    excitation in case of horizontal pulsing confirms the excitation
    of the $7 Q_x$ resonance. For \tenthtp, there is in contrast no
    significant difference between H, V, or H+V
    (Fig.~\ref{fig:fma:10}).}
  \label{fig:patternfma}
\end{figure*}

@@

As an example, the simulation results for the 2016 experiment are
shown in Fig.~\ref{fig:patternsim}.  A clear dependence of both losses
and emittances on the pattern can be seen.  The largest losses are
predicted for 3rd-, 7th- and \tenthtp, and for a uniform random
excitation. Because the bunch length decreases with the number of lost
particles, losses are attributed to off-momentum particles hitting the
transverse aperture. Significant emittance growth is visible only for
7th- and \tenthtp\ and for the random excitation. Compared to any
resonant excitation, the random excitation shows by far the strongest
effect.

The effects of 7th- and \tenthtp\ are also observed in the absence of
machine lattice errors. In this case, the only sources of nonlinearity
are sextupoles and octupoles~\cite{md_sim_hel_res_ex_fitterer},
suggesting that these nonlinearities are responsible for the
pronounced beam sensitivity. The driven resonances are revealed by the
frequency-map analysis, shown in Fig.~\ref{fig:patternfma}. The
$7 Q_x$ resonance is excited by the \seventhtp, and the $10 Q_x$ and
$10 Q_y$ resonances by the \tenthtp. As octupoles can only drive even
resonances, the sources of the $7 Q_x$ resonances are the sextupoles,
while the octupoles generate the tune footprint. The other pulsing
patterns do not exhibit any increase in losses or emittance growth
without magnetic errors. Their effect can be thus attributed to an
interaction of the excitation with the lattice errors, implying also a
sensitivity to the chosen random seed in simulations and to the
specific lattice in experiments.

@@

\subsection{Pulsing every 10th turn}
\label{sec:simex10}

\begin{figure*}
  \begin{tabular}{ccc}
    \includegraphics[width=\thirdwidth]{2016_bunch_intensity_v10th_no_damper_avg.png}&
    \includegraphics[width=\thirdwidth]{2016_emith_avg_rel_v10th_no_damper.png}&
    \includegraphics[width=\thirdwidth]{2016_emitv_avg_rel_v10th_no_damper.png}\\
  \end{tabular}
  \caption{Summary of the 2016 experiments on \tenthtp\ in the
    vertical plane: losses (left), horizontal emittances (middle) and
    vertical emittances (right), relative to their initial values. The
    transverse damping system was not active in this case. The 3
    excitation periods are labeled in black according to the value of
    the maximum excitation amplitude
    $A_{\mathrm{max}} = 5 \, \Delta A$: no excitation, 48~nrad or
    96~nrad. The 4~bunches experiencing the same excitation amplitude
    $n \, \Delta A$ ($n = 0, \ldots, 5$) are grouped by color. The
    data are averaged over the 4~bunches, with the envelope
    representing the standard deviation. The area with a blue
    background highlights qualitatively the fast adjustment period of
    the beam distribution, transitioning into a new equilibrium state
    (indicated by the gray background).}
  \label{fig:10thexp}
\end{figure*}

@@

An insight on how the change in tune entails a change in driving
resonances is provided by frequency-map analysis.
Figure~\ref{fig:patternfma} (bottom left and bottom right plots),
which was described previously, shows a strong increase in diffusion
for \seventhtp\ in the horizontal plane around the $7Q_x$ resonances,
and only small changes for pulsing in the vertical plane. In the case
of the 2017 optics and tune, both the $7Q_x$ and the $7Q_y$ resonances
cross the tune footprint (Fig.~\ref{fig:7th2017fma}, top left). Both
of these resonances are excited by \seventhtp, as one can see from the
increase in tune jitter around the corresponding resonance lines for
each excitation plane separately, and thus also for their combined
action (Fig.~\ref{fig:7th2017fma}, bottom left, bottom right, and top
right plots). Although much weaker, an increase in tune diffusion is
also observed around the $7Q_x + 7Q_y$ line when pulsing in the
vertical plane or in both planes (diagonal yellow-green line in
Fig.~\ref{fig:7th2017fma}, top right and bottom right plots). This
suggests an excitation of the 14th-order resonance. The reason why
this phenomenon is much less pronounced for horizontal excitations is
not understood.
